Developing Observational Health Research Capacity in the Julia Ecosystem

### 2023-12-26 paper.tex
Finished full review of JuliaCon paper

% TODO: Reference code repository
Using the JuliaHealth tools developed, we enumerate an example observational health study workflow that would be used in practice.\footnotemark
For this paper, we used two datasets: a synthetic patient dataset called \textit{Eunomia} and the MIMIC III patient dataset which had been converted into the OMOP CDM to illustrate the workflow provided by the Observational Health sub-ecosystem.
Reproducing the workflow using the \textit{Eunomia} dataset is made readily available at the repository here: \cite{schuemieEunomia2023}

@@

@@

\subsection{Bridging between OHDSI and JuliaHealth Communities}

@@

After having a strep throat phenotype definition defined using ATLAS, we can then execute the resulting computable phenotype definition against the \textit{Eunomia} database.

\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
using JSON3 
using OHDSICohortExpressions

@@

A quick query of the cohort table (where the resulting subjects were stored) shows $\approx 1600$ from the dataset match this criteria.

@@

\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
C = GetCohortSubjects(1, conn)
rename!(C, :subject_id => :person_id)

@@

@@

Using this analytical approach, tabulation can be quickly conducted across patients.
Additionally, throughout the process, one can perform relevant analysis methods.
Here, we'll conduct a crude prevalence calculation across the patient cohort.

\begin{table}[!ht]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|}
    \hline
        Race & Gender & Age Group & Cohort & Total & Prev. (\%) \\ \hline
        B/AA & F & 60 - 64 & 13 & 21 & 61.90 \\ \hline
        W & F & 50 - 54 & 59 & 111 & 53.15 \\ \hline
        W & F & 55 - 59 & 62 & 97 & 63.91 \\ \hline
        A & M & 80 - 84 & 6 & 10 & 60.00 \\ \hline
        UNK & F & 60 - 64 & 29 & 39 & 74.35 \\ \hline
        B/AA & M & 65 - 69 & 12 & 18 & 66.66 \\ \hline
    \end{tabular}
\end{table}

@@

As seen in Table 1 and Table 2, informative results can be rapidly generated by this workflow relevant to observational health research.
Although the values shown in Table 2 are synthetic and the crude prevalence values themselves are meaningless, we can take a this approach and apply it directly to the MIMIC III dataset.
In Table 3, we can calculate across this dataset meaningful statistics that could later be further analyzed and drive further questioning (such as ...).

@@

    1. Modularization of code and chaining
    2. Distributed computing

\subsubsection{Modularization of Code and Chaining}

An exciting emergent aspect of the JuliaHealth ecosystem that came about as a result of developing OMOPCDMCohortCreator.jl was the prospect of radical modularization and chaining of package functionalities together.
For example, taking our workflow example from the earlier section where we characterized patients by race, gender, and age group, we can do a few "tricks" to modularize these characterization functions as follows:

We can leverage a unique property of Julia in that Julia supports partial evaluation of functions (also known as currying) to simplify repeated function calls to the same function over and over again.
First, we'll take our original functions and fix a connection object to each of our cofactor functions.
This simplifies having to continuously pass forward a connection object and opens up the ability to chain together these functions simply:

\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
import Base:
  Fix2

@@

NOTE: The "C" prefix for each of our curried functions denotes that it is curried.

Second, as OMOPCDMCohortCreator.jl is built on top of DataFrames.jl and one has the option of creating a DataFrame with the result from every OCC function, we can use the package, Chain.jl.
This package abstracts away some of the explicit functionalities of DataFrames.jl but instead gives much easier ways to define a flow of functions:

\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
import Chain:
  @chain

@@

\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
CharacterizePatients(patients) = @chain patients begin
    _[:, DF.Not(:person_id)]
    DF.groupby(_, names(_))
    DF.combine(_, DF.nrow => :count)
    occ.ExecuteAudit
end
\end{minted}

@@

In this way, future researchers using tools such as OMOPCDMCohortCreator.jl can create these more convenient abstractions.
In fact, it could be a point of exploration in the future to build on top of OMOPCDMCohortCreator.jl even more abstracted and reusable functionalities to simplify ease of use while also enabling more readily auditable functionality.
Finally, as these functions have now been fully modularized, these can now be readily distributed using Julia's distributed computing methods:

\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
import Distributed:
  @distributed

@@

One crucial aspect of the OHS is that it is unavoidable to work with databases that contain patient health information.
For better or worse, the tool that is used to best interact with these databases are SQL dialects that loosely follow consistently the ANSI SQL guidance.
To accomodate for this, several of the core tools developed in this sub-ecosystem generate SQL that can be used in the case where Julia may not be available in a given environment housing patient health data.
As an example, \textit{OMOPCDMCohortCreator.jl} functions build upon \textit{FunSQL.jl} allowing one to use a function such `GetPatientAgeGroup` without a connection object which would result in SQL.

\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
  GetPatientAgeGroup([1]; age_groupings = [[0, 40], [41, 80]])
\end{minted}

\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
  SELECT
    "PERSON_2"."person_id",
    (CASE 
      WHEN ("PERSON_2"."age" < 41) 
        THEN '0 - 40' 
      WHEN ("PERSON_2"."age" < 81) 
        THEN '41 - 80' END) AS "age_group"
  FROM (
    SELECT
      "PERSON_1"."person_id",
      (2023 - "PERSON_1"."year_of_birth") AS "age"
    FROM "PERSON" AS "PERSON_1"
    WHERE ("PERSON_1"."person_id" IN (1))
  ) AS "PERSON_2"
\end{minted}

\subsection{Strong Composition within JuliaHealth}

@@

As shown in this work, there is a rich opportunity for the JuliaHealth organization to become a strong presence within observational health research. 
JuliaHealth is now sufficiently at a stage of maturity where potential future researchers can leverage JuliaHealth's tools to conduct further future research or build upon existing architecture to target specific needs.
Some potential future directions that can be taken are as follows.

Due to the strong composition that exists within the JuliaHealth and broader Julia ecosystem, solutions may already exist for solving various health informatics related research.
However, what has consistently been a problem within the Julia ecosystem is the lack of sufficient documentation dedicated to describing these novel compositions and their application to various endeavors.
Future efforts can be dedicated to taking existing tools and reframing them in a JuliaHealth context to assist newcomers to Julia and JuliaHealth in determining how to explore and address problems that they are interested in.

Additionally, due to the composable and modular nature of the OHS, research analyses and visualizations can be decoupled and interchanged with one another.
Whereas in other ecosystems that make technologies for tools such as dashboards tightly coupled to analytics regimens, the subecosystem is mature enough to enable new users and developers to create analytics tools that can be separately woven into more graphical user interfaces.
For example, treatment pathways is a novel observational health approach to understand and visualize the care a patient receives within a care setting.
Only a few tools exist to visualize such hidden patient narratives but with the JuliaHealth ecosystem, the possibility of making tools that can interrogate patient care pathways more deeply could be fully realized.

Furthermore, again exploiting the composition within the JuliaHealth ecosystem, as mentioned in the context of JuliaHealth and machine learning applications, future efforts could focus on developing fairness and bias auditing tools in patient datasets.
These tools could consider factors commonly explored in health equity literature like data completeness, demographic variables, and fairness algorithms.

Moreover, there's potential to advance clinical informatics by developing a flexible clinical type system, supporting computable phenotype definitions and medical ontologies. The ecosystem can also benefit from tools to audit the fairness of phenotype definitions. Finally, as the landscape of data expands, integrating non-traditional sources like social media and environmental data into standardized patient records will be a critical frontier. In these ways, JuliaHealth is primed to drive innovation in observational health informatics, pushing the boundaries of data-driven healthcare research.

@@

In conclusion, this paper has demonstrated the capabilities of the JuliaHealth OHS and its potential to further investigations in observational health research.
Although it is an ecosystem that is relatively new, its robustness allows for novel investigations to be pursued within this domain of research. 
Additionally, due to the strongly compositional nature of the Julia ecosystem, these JuliaHealth tools can readily compose with tooling to support in-depth statistical analyses, mathematical modeling, and visualization capacities.

The deep dive into the JuliaHealth ecosystem unravels its arsenal of tools tailored for observational health research, meticulously designed to navigate the complexities of handling OMOP CDM data. Notably, the sub-ecosystem within JuliaHealth presents mature, internationally federated research capabilities, building upon existing innovations like OHDSI's HADES, while offering novel tools such as OMOPCDMCohortCreator.jl.
In considering future directions, the paper posits the promising avenues for JuliaHealth, envisioning expanded documentation, the development of visualization tools, and integration of machine learning applications for fairness auditing. The ecosystem's potential to revolutionize clinical informatics by integrating non-traditional data sources, auditing phenotype definitions, and fostering innovation in data-driven healthcare research is paramount.
In essence, the JuliaHealth ecosystem stands as a catalyst for transformative observational health research, poised to push the boundaries of innovation, collaboration, and standardization in the pursuit of better healthcare outcomes. As it continues to evolve, JuliaHealth emerges as a beacon of progress in the dynamic landscape of health informatics.

@@

  \item Dr. Jon Duke for his early support
  \item Kristin Kostka for being an early champion in trying novel ideas
  \item Dilum Aluthge for helping this work find a home in JuliaHealth
  \item Adam Black for thinking through potential approaches
  \item The Roux Institute's OHDSI Center for their encouragement
  \item The JuliaHealth community and contributors for fearlessly trying to push boundaries on what observational health informatics can truly be

@@

>>>>

% TODO: Reference code repository
Using the JuliaHealth tools developed, we enumerate an example observational health study workflow that would be used in practice.\footnotemark
% TODO: Enumerate these two datasets
% EXPLAIN: In enumeration talk about what Eunomia is 
% EXPLAIN: In enumeration discuss MIMIC III
% TODO: Cite MIMIC III, MIT LCP, and OMOP-ification process
For this paper, we used two datasets: a synthetic patient dataset called \textit{Eunomia} and the MIMIC III patient dataset which had been converted into the OMOP CDM to illustrate the workflow provided by the Observational Health sub-ecosystem.
% TODO: Mention in footnote explicitly that reproducing workflow in Eunomia is possible
% TODO: Mention in footnote that explicitly reproducing workflow in MIMIC III is possible but not readily accessible
Reproducing the workflow using the \textit{Eunomia} dataset is made readily available at the repository here: \cite{schuemieEunomia2023}

@@

% EXPLAIN: What is this doing here?
% TODO: Update DBConnector
% TODO: Maybe remove package imports and mention them within the caption instead

@@

% TODO: Review if this subsection is needed
\subsection{Bridging between OHDSI and JuliaHealth Communities}

@@

% EXPLAIN: What is this doing here?
After having a strep throat phenotype definition defined using ATLAS, we can then execute the resulting computable phenotype definition against the \textit{Eunomia} database.

% TODO: Add figure, caption, and label details
% TODO: Maybe remove package imports and mention them within the caption instead
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
using JSON3 
using OHDSICohortExpressions

@@

% TODO: Move this to a caption
A quick query of the cohort table (where the resulting subjects were stored) shows $\approx 1600$ from the dataset match this criteria.

@@

% EXPLAIN: What is this section doing?

% TODO: Add figure, caption, and label details
% TODO: Maybe remove package imports and mention them within the caption instead
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
C = GetCohortSubjects(1, conn)
rename!(C, :subject_id => :person_id)

@@

% EXPLAIN: What is this section doing?

% TODO: Add figure, caption, and label details
% TODO: Maybe remove package imports and mention them within the caption instead

@@

Using this analytical approach, tabulation can be quickly conducted across patients.

% EXPLAIN: Move this to the caption of the table
Additionally, throughout the process, one can perform relevant analysis methods.
Here, we'll conduct a crude prevalence calculation across the patient cohort.

% EXPLAIN: What is this section doing?

% TODO: Add figure, caption, and label details
\begin{table}[!ht]
    \centering
    \begin{tabular}{|l|l|l|l|l|l|}
    \hline
        Race & Gender & Age Group & Cohort & Total & Prev. (\%) \\ \hline
        B/AA & F & 60 - 64 & 13 & 21 & 61.90 \\ \hline
        W & F & 50 - 54 & 59 & 111 & 53.15 \\ \hline
        W & F & 55 - 59 & 62 & 97 & 63.91 \\ \hline
        A & M & 80 - 84 & 6 & 10 & 60.00 \\ \hline
        UNK & F & 60 - 64 & 29 & 39 & 74.35 \\ \hline
        B/AA & M & 65 - 69 & 12 & 18 & 66.66 \\ \hline
    \end{tabular}
\end{table}

@@

% TODO: Create table for MIMIC III results
% TODO: Create reference to Eunomia table
As seen in Table 1, informative results can be rapidly generated by this workflow relevant to observational health research.
Although the values shown in Table 2 are synthetic and the crude prevalence values themselves are meaningless, we can take a this approach and apply it directly to the MIMIC III dataset.
In Table 3, we can calculate across this dataset meaningful statistics that could later be further analyzed and drive further questioning (such as ...).

@@

% EXPLAIN: What is the overview of this section?
% EXPLAIN: What does each topic do?
% TODO: Convert this to an enumeration

    1. Modularization of code and chaining
    2. Distributed computing
    3. Automatic generation of SQL from health packages

% TODO: Rename section
\subsubsection{Modularization of Code and Chaining}

% TODO: Make this more explicit about utilizing Julia's composition
An exciting emergent aspect of the JuliaHealth ecosystem that came about as a result of developing OMOPCDMCohortCreator.jl was the prospect of radical modularization and chaining of package functionalities together.
% TODO: Tie into currying more directly
% TODO: Check if I am mentioning currying correctly
We can leverage a unique property of Julia in that Julia supports partial evaluation of functions (also known as currying) to simplify repeated function calls to the same function over and over again.
% TODO: Rewrite this part in light of the previous sections 
For example, taking our workflow example from the earlier section where we characterized patients by race, gender, and age group, we can do a few "tricks" to modularize these characterization functions as follows:

% TODO: Rework this section further
First, we'll take our original functions and fix a connection object to each of our cofactor functions.
This simplifies having to continuously pass forward a connection object and opens up the ability to chain together these functions simply:

% TODO: Maybe move this into the tutorial code
% TODO: Add figure, caption, and label details
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
import Base:
  Fix2

@@

% TODO: Add this to previous code caption
NOTE: The "C" prefix for each of our curried functions denotes that it is curried.

% EXPLAIN: Refactor this paragraph better or put most of it into the caption for the next code block
Second, as OMOPCDMCohortCreator.jl is built on top of DataFrames.jl and one has the option of creating a DataFrame with the result from every OCC function, we can use the package, Chain.jl.
This package abstracts away some of the explicit functionalities of DataFrames.jl but instead gives much easier ways to define a flow of functions:

% TODO: Do we even need chaining here or could we use direct composition of functions?
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
import Chain:
  @chain

@@

% TODO: Can this be simplified?
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
CharacterizePatients(patients) = @chain patients begin
    _[:, DF.Not(:person_id)]
    DF.groupby(_, names(_))
    DF.combine(_, DF.nrow => :count)
    occ.ExecuteAudit
end
\end{minted}

@@

In this way, future researchers using tools such as OMOPCDMCohortCreator.jl can create these more convenient abstractions.
% TODO: Move this into the future discussion section
In fact, it could be a point of exploration in the future to build on top of OMOPCDMCohortCreator.jl even more abstracted and reusable functionalities to simplify ease of use while also enabling more readily auditable functionality.

% TODO: Move this into its own section
Finally, as these functions have now been fully modularized, these can now be readily distributed using Julia's distributed computing methods:

% TODO: Redo example; I do not feel it is good
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
import Distributed:
  @distributed

@@

% TODO: Rephrase this sentence
One crucial aspect of the OHS is that it is unavoidable to work with databases that contain patient health information.
% TODO: Rephrase this sentence better
% TODO: Cite ANSI SQL
For better or worse, the tool that is used to best interact with these databases are SQL dialects that loosely follow consistently the ANSI SQL guidance.
% EXPLAIN: Highlight this more
% EXPLAIN: In what scenarios might using Julia may not be possible
To accomodate for this, several of the core tools developed in this sub-ecosystem generate SQL that can be used in the case where Julia may not be available in a given environment housing patient health data.
% EXPLAIN: Introduce this better
As an example, \textit{OMOPCDMCohortCreator.jl} functions build upon \textit{FunSQL.jl} allowing one to use a function such `GetPatientAgeGroup` without a connection object which would result in SQL.

% TODO: Add figure, caption, and label details
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
  GetPatientAgeGroup([1]; age_groupings = [[0, 40], [41, 80]])
\end{minted}

% TODO: Add figure, caption, and label details
\begin{minted}[breaklines,escapeinside=||,mathescape=true, numbersep=3pt, gobble=2, frame=lines, fontsize=\scriptsize, framesep=2mm]{julia}
  SELECT
    "PERSON_2"."person_id",
    (CASE 
      WHEN ("PERSON_2"."age" < 41) 
        THEN '0 - 40' 
      WHEN ("PERSON_2"."age" < 81) 
        THEN '41 - 80' END) AS "age_group"
  FROM (
    SELECT
      "PERSON_1"."person_id",
      (2023 - "PERSON_1"."year_of_birth") AS "age"
    FROM "PERSON" AS "PERSON_1"
    WHERE ("PERSON_1"."person_id" IN (1))
  ) AS "PERSON_2"
\end{minted}

% TODO: Combine this section with the future directions section
\subsection{Strong Composition within JuliaHealth}

@@

% TODO: Rephrase this introductory paragraph
JuliaHealth is now sufficiently at a stage of maturity where potential future researchers can leverage JuliaHealth's tools to conduct further future research or build upon existing architecture to target specific needs.
Some potential future directions that can be taken are as follows.

% TODO: This is where to introduce the paradox of composition
Due to the strong composition that exists within the JuliaHealth and broader Julia ecosystem, solutions may already exist for solving various health informatics related research.
However, what has consistently been a problem within the Julia ecosystem is the lack of sufficient documentation dedicated to describing these novel compositions and their application to various endeavors.
Future efforts can be dedicated to taking existing tools and reframing them in a JuliaHealth context to assist newcomers to Julia and JuliaHealth in determining how to explore and address problems that they are interested in.

% TODO: Combine this paragraph with the previous paragraph
% TODO: Shorten this sentiment
Additionally, due to the composable and modular nature of the OHS, research analyses and visualizations can be decoupled and interchanged with one another.
Whereas in other ecosystems that make technologies for tools such as dashboards tightly coupled to analytics regimens, the subecosystem is mature enough to enable new users and developers to create analytics tools that can be separately woven into more graphical user interfaces.
For example, treatment pathways is a novel observational health approach to understand and visualize the care a patient receives within a care setting.
Only a few tools exist to visualize such hidden patient narratives but with the JuliaHealth ecosystem, the possibility of making tools that can interrogate patient care pathways more deeply could be fully realized.

% TODO: Make this more about the flexible nature of the JuliaHealth organization
Furthermore, again exploiting the composition within the JuliaHealth ecosystem, as mentioned in the context of JuliaHealth and machine learning applications, future efforts could focus on developing fairness and bias auditing tools in patient datasets.
These tools could consider factors commonly explored in health equity literature like data completeness, demographic variables, and fairness algorithms.

@@

In conclusion, this paper has demonstrated the capabilities of the JuliaHealth OHS and its potential to further investigations in observational health research.
Although it is an ecosystem that is relatively new, its robustness allows for novel investigations to be pursued within this domain of research. 
% TODO: Mention how this allows JuliaHealth to solve more problems than what is present within JuliaHealth
Additionally, due to the strongly compositional nature of the Julia ecosystem, these JuliaHealth tools can readily compose with tooling to support in-depth statistical analyses, mathematical modeling, and visualization capacities.

% TODO: Rework this
In essence, the JuliaHealth ecosystem stands as a catalyst for transformative observational health research, poised to push the boundaries of innovation, collaboration, and standardization in the pursuit of better healthcare outcomes. As it continues to evolve, JuliaHealth emerges as a beacon of progress in the dynamic landscape of health informatics.

@@

  \item Dr. Jon Duke for his early support
  \item Kristin Kostka for being an early champion in trying novel ideas
  \item The JuliaHealth community and contributors for fearlessly trying to push boundaries on what observational health informatics can truly be
  \item Dilum Aluthge for helping this work find a home in JuliaHealth
  \item Adam Black for thinking through potential approaches
  \item Clark Evans and Kirill Simonov for being the pioneers of this work in Julia
  \item The Roux Institute's OHDSI Center for their encouragement

@@

% TODO: Review and remove appendix
